Pleiotropy of expression quantitative trait loci increases with the number of regulated genes, active cell types and tissues and bound transcription factors

### 2022-12-18 ms/ms03_results.tex
ENH PMIDs in disease matrix

To better understand the role of gene regulatory variants for variant pleiotropy,
I developed a pipeline to systematically colocalize variants from genome-wide associations
studies (GWAS) and expression quantitative trait loci (eQTL) (Section Methods).
%
127 eQTL studies were downloaded from the EBI eQTL catalogue, which aims to provide uniformly processed eQTLs in many
tissues and cell types \citep{2021.Alasoo.Kerimov} (Supplementary table 1).

GWAS traits were selected from the IEU OpenGWAS database based on four criteria \citep{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most
samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological
conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS studies with at least 10,000 subjects, 2,000 controls and 2,000 cases.
%
These filters resulted in 418 GWAS traits (Supplementary table 2).
%
Among these studies, there were 10,627 clumped leading variants with a p-value below 5e-8 from 335 GWAS (Supplementary file).
% sqlite
% select count(distinct gwas_id) from tophits; 335
% select count(*) from (select DISTINCT chrom, pos, rsid, nea, ea from tophits) limit 50; 10627
% Supplementary file is db.sqlite that can go to OSF

My pipeline uncovered eQTL colocalizations for 3,849 or 36$\%$ of the leading GWAS variants with 2,348 eQTL genes and 127 eQTL samples.
% sqlite
% select count(distinct tophits.rsid) from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.`PP.H4.abf`>=0.8; 3849
% alternative: select count(distinct tophits.rsid) from tophits, coloc where coloc.rsid=tophits.rsid and coloc.`PP.H4.abf`>=0.8; 3849
% select count(distinct coloc.egene) from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.`PP.H4.abf`>=0.8
% select count(distinct coloc.eqtl_id) from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.`PP.H4.abf`>=0.8

The largest classes of diseases by number of leading GWAS variants are autoimmune diseases (2107), cancer of breast (1976) and allergy (1321)
%
In these diseases, the percentages of variants with a colocalized eQTL are 37\%, 21\% and 42\% in these classe, respectively (Supplementary table ST3).
%
%/* count tophits.rsid group by gwas_annot.gwas_class where coloc.`PP.H4.abf`>=0.8*/
%select gwas_annot.gwas_class, count(distinct tophits.rsid) from tophits, coloc, gwas_annot where gwas_annot.gwas_id=tophits.gwas_id and coloc.rsid=tophits.rsid and coloc.`PP.H4.abf`>=0.8 GROUP by gwas_annot.gwas_class;
%/* count tophits.rsid group by gwas_annot.gwas_class*/
%select gwas_annot.gwas_class, count(distinct tophits.rsid) from tophits, gwas_annot where gwas_annot.gwas_id=tophits.gwas_id GROUP by gwas_annot.gwas_class order by gwas_annot.gwas_class;
%
These results agree with previous work for autoimmune diseases \citep{2021.Li.Mu}.

This pipeline resulted in 103,551 variants from 246 GWAS that colocalized (PP.H4.abf$\geq$0.8) with at least one eQTL.
%select count(*) from (select DISTINCT gwas_id FROM coloc where `PP.H4.abf`>=0.8);
%select count(*) from (select DISTINCT rsid FROM coloc where `PP.H4.abf`>=0.8);
%
I have also developed a web application where I expose 116,736 variants from 246 GWAS that colocalize with at least one eQTL at a lower threshold value PP.H4.abf$\geq$0.5.
% select count(*) from (select DISTINCT rsid FROM coloc where `PP.H4.abf`>=0.5);
% select count(*) from (select DISTINCT gwas_id FROM coloc where `PP.H4.abf`>=0.5);

%\subsection*{Pleiotropic variants colocalized with eQTLs}

\subsection*{Traits classification based on the colocalized eQTL effect size}

I manually assigned the 418 GWAS to 96 classes to aggregate identical or similar traits (Supplementary table 2).
%
Then I verified whether GWAS traits cluster coherently within these classes.
%
I computed distances between GWAS traits based on the Spearman correlation of the colocalized eQTL beta for every target gene, cell type and tissue.
%
I plot only classes with more than a minimum number of GWAS traits (Fig. \ref{fig:gwas_distance}).
%
GWAS traits cluster together in autoimmune diseases and circulatory system diseases.
%
Due to the large number of breast cancer GWAS, we have created a separated class from cancer.
Nevertheless, we can observe weaker correlations between the breast cancer class and other types of cancers.

@@

Very pleiotropic regions with 5 trait classes or more make around 0.6 Mb of the genome,
regions with 4 GWAS classes make an additional 1 Mb and regions with 3 trait classes make an additional 1.5 Mb (Fig. \ref{fig:pleiotropy_region_distribution}b).
%
Altogether, pleiotropic regions with 3 or more trait classes make around 3 Mb of the genome (Fig. \ref{fig:pleiotropy_region_distribution}b).

>>>>

To annotate traits from genome-wide associations studies (GWAS) with genes and tissues,
I developed a pipeline that systematically colocalizes variants from GWAS and expression quantitative trait loci (eQTLs) (Section Methods).
%
% TODO Supplementary table 1 for newest pipeline
127 eQTL studies were downloaded from the EBI eQTL catalogue, which aims to provide uniformly processed eQTLs in many
tissues and cell types \citep{2021.Alasoo.Kerimov} (Supplementary table 1).

GWAS traits were selected from the IEU OpenGWAS database based on four criteria \citep{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most
samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological
conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS with at least 10,000 subjects, 2,000 controls and 2,000 cases.
%
% TODO Supplementary table 2 for newest pipeline
These filters resulted in 418 GWAS traits (Supplementary table 2).
%
Among these studies, there were 8,580 clumped leading variants with a p-value below 5e-8 from 284 GWAS (Supplementary file).
% select count(distinct gwas_id) from tophits; 284
% select count(*) from (select DISTINCT chrom, pos, rsid, ea from tophits) as sel; 8,580
% tophits goes to OSF
% colocpleio goes to OSF

My pipeline uncovered eQTL colocalizations for 5,484 or 64$\%$ of the leading GWAS variants with 7,040 eQTL genes and 127 eQTL biosamples.
% select count(*) from (select distinct tophits.rsid from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75) as sel; 5484
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75) as sel;  7040
% select count(*) from (select distinct eqtl_id from coloc where pp_h4_abf>=0.75) as sel;  127

The largest classes of diseases by number of leading GWAS variants are autoimmune diseases (2107), cancer of breast (1976) and allergy (1321)
%
In these diseases, the percentages of variants with a colocalized eQTL are 37\%, 21\% and 42\% in these classe, respectively (Supplementary table ST3).
% TODO
% Number of tophits rsid per GWAS class
% select count(distinct tophits.rsid),gwas_class from gwas_annot, tophits where tophits.gwas_id=gwas_annot.gwas_id group by gwas_class order by count desc;
% Number of tophits rsid per GWAS class that is also preset in coloc table
% select count(distinct tophits.rsid),gwas_class from gwas_annot, tophits, coloc where tophits.gwas_id=gwas_annot.gwas_id and tophits.rsid=coloc.rsid group by gwas_class order by count desc;
% OLD SQLITE
% select count(distinct tophits.rsid),gwas_class from gwas_annot join tophits on tophits.gwas_id=gwas_annot.gwas_id group by gwas_class order by count desc limit 5;
%/* count tophits.rsid group by gwas_annot.gwas_class where coloc.`PP.H4.abf`>=0.8*/
%select gwas_annot.gwas_class, count(distinct tophits.rsid) from tophits, coloc, gwas_annot where gwas_annot.gwas_id=tophits.gwas_id and coloc.rsid=tophits.rsid and coloc.`PP.H4.abf`>=0.8 GROUP by gwas_annot.gwas_class;
%/* count tophits.rsid group by gwas_annot.gwas_class*/
%select gwas_annot.gwas_class, count(distinct tophits.rsid) from tophits, gwas_annot where gwas_annot.gwas_id=tophits.gwas_id GROUP by gwas_annot.gwas_class order by gwas_annot.gwas_class;

These results are consistent with previous colocalization analysis that found explained percentages of work for autoimmune diseases \citep{2021.Li.Mu}.....
% TODO
% Discuss 25% in Chun et al 2017, Nature Genetics
% Discuss 41% (eQTL + sQTLs) in Mu et al 2021, Genome Biology
% Discuss 23% in Chen et al 2016, Cell. Genetic Drivers of Epigenetic and Transcriptional Variation in Human Immune Cells

This pipeline resulted in 103,551 variants from 246 GWAS that colocalized (PP.H4.abf$\geq$0.8) with at least one eQTL.
% TODO
%select count(coloc.rsid) from (select DISTINCT gwas_id FROM coloc where pp_h4_abf>=0.8);
%select count(*) from (select distinct rsid from coloc where pp_h4_abf>=0.8) as sel;
%
This might be a useful resource to annotate GWAS variants with genes and biosamples.
Therefore I have developed a web application where I expose 116,736??? variants from 246??? GWAS that colocalize with at least one eQTL at a threshold value PP.H4.abf$\geq$0.75.
% TODO
% select count(*) from (select DISTINCT chrom,pos38,alt FROM colocweb) as sel;
% select count(*) from (select DISTINCT gwas_id FROM coloc) as sel;

% Work with PP.H4.abf`>=0.75 and SNP.PP.H4>=0.5
\subsection*{Traits classification based on the colocalized eQTL effect size}

In the previous colocalization analysis, we found many eQTLs that are associated to very different traits.
%
To systematically identify pleiotropic eQTLs, I manually annotated the GWAS traits in different classes.

% TODO Check if I aggregate different coloc windows or not.
I also computed trait distances based on the spearman correlation of eQTL effect size for distinct eQTLs given by
distinct chromosome, position, alternative allele, eQTL gene and eQTL biosample.
%
From this point on I defined colocalization with parameters PP.H4.abf>=0.75 and SNP.PP.H4>=0.5.
%
The argument for selecting a cutoff SNP.PP.H4>=0.5 is given below.
%
I plotted these trait distances as a hierarchically-clustered heatmap.
%
I plotted only trait classes with more than a minimum number of traits (Fig. \ref{fig:gwas_distance}).
%
Due to the large number of datasets, I have create a separate class for some cancers such as breast and skin.
%
This analysis shows a coherent clustering of traits belonging to autoimmune, circulatory system diseases and cancers (Fig. \ref{fig:gwas_distance}).
%

Nevertheless, we can observe weaker correlations between the breast cancer class and other types of cancers.

% TODO create supplementary table 2
I manually assigned the 418 GWAS to 96 classes to aggregate identical or similar traits (Supplementary table 2).
%
Then I verified whether GWAS traits cluster coherently within these classes.
%
I computed distances between GWAS traits based on the Spearman correlation of the colocalized eQTL beta for every target gene, cell type and tissue.
%
I plot only classes with more than a minimum number of GWAS traits (Fig. \ref{fig:gwas_distance}).
%
GWAS traits cluster together in autoimmune diseases and circulatory system diseases.
%
Due to the large number of breast cancer GWAS, we have created a separated class from cancer.
Nevertheless, we can observe weaker correlations between the breast cancer class and other types of cancers.

@@

Very pleiotropic regions with 5 trait classes or more make around 0.6 Mb of the genome,
regions with 4 GWAS classes make an additional 1 Mb and regions with 3 trait classes make an additional 1.5 Mb (Fig. \ref{fig:pleiotropy_region_distribution}b).
%
Altogether, pleiotropic regions with 3 or Routinesmore trait classes make around 3 Mb of the genome (Fig. \ref{fig:pleiotropy_region_distribution}b).
